\chapter*{Acknowledgment}
First of all, I thank my supervisors, Pr. Engelbert Mephu Nguifo and Dr.  Philippe Vaslin for following my work since Master 2. For their high availability, their support and for all the educational exchanges we had. Beyond the thesis work, they are admirable people whose qualities positively influence my person. Thank you very much for all.

I thank LIMOS, which has allowed me to carry out these years of work in a stimulating environment and for having placed at my disposal everything necessary for the realization of this work. I thank  the Ministry of Higher Education, Research and Innovation  for the support it gives to the research and the researcher in general, and for having financed my thesis work. 

I thank the members of my thesis jury for having accepted to take part in this jury. I thank them for the time they devoted to reading the manuscript, for their questions and their remarks which will allow me to improve the quality of this work.

A special thanks to my teachers in general, to those in particular with whom I exchanged during this work; I think of Libo Ren and Violaine Antoine. I also thank Sévérine Miginiac and Béatrice Bourdieu for their kindness and their availability.

I spent pleasant moments with my doctoral colleagues from LIMOS whom I would like to thank. The list is too long to mention them all, I will mention Damien Lamy with whom I did the master, Maxime Chassaing and Benjamin Vincent who welcomed me the first year of my thesis, Benjamin Dalmas for our informative discussions and for his good humour, Angeline Plaud, Nestor Koueya, Romain, Jerry Lonlac, my colleagues and friends from MINERS, Nina Bekono and Raissa Saleu my little sisters. 
A special thank you also to my colleagues in the office with whom we made the world over coffee or a meal, I think of Jinhua Zhao, Jocelyn de Goer, Suan Tay, Danh Nguyen, Ismaela Seck, Rachid Jijiane, Jean Conier, Alexis Cornet, Carlos Cepeda.

Finally, I thank my family for their support, encouragement and wise advice. I think of the Siyou’s Family, Njomgang’s Family, Ngainsi’s Family  and also my whole big family. I especially thank my beloved Lucia and our son Luka for all the love and affection they bring to me. 
